Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-27 manuscript/paper.tex
Polish UI section

\subsection{User Interface}
\label{subsec:User_Interface}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
    \caption{\textbf{A view of ARG edges in \texttt{tsbrowse}.}
    A screenshot of the Edges tab for a 5 Mb sequence simulated with selection coefficient, \textit{s = 0.25}, and beneficial allele situated at \textit{x = 2.5 Mb}. The plot on the left represents each edge as a horizontal line along the sequence (X-axis). The time of parent node is shown on the Y-axis. The persistence of edges across various dimensions is shown through histograms on the right: top, distribution of genomic spans; middle, distribution of time spans; bottom, edge area, which is the product of genomic and time spans.}

    \label{fig:Figure_2}
\end{figure}
% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
% visualisation pipelines even for large datasets. A key challenge with
% large-scale data visualisation is to maintain a responsive user interface.
% \texttt{Datashader} aggregates data and renders only the necessary data points
% at a given zoom level, thereby minimising the computational load on the
% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
% with individual data points, allowing users to gain deeper insights from the
% data. The \texttt{serve} step accepts an optional input file containing
% sequence annotations. This allows users to overlay contextual information about
% genes or other sequence features. 
The user interface is presented as a dashboard, with pages to
describe mutations, nodes, edges and trees in the input ARG. Figure 2
demonstrates the user interface using the Edges page as an example. The plot on
the left is an overview of all 33,929 edges (parent-child relationships between
node-pairs; see \hyperref[subsec:Data_Model]{Data Model}) in a simulated ARG
\hyperref[subsec:sweep_simulation]{described in Supplementary Methods}; each
edge is depicted as a horizontal line connecting the genomic coordinates of the
parent and child nodes on the x-axis. The y-axis denotes the age of either
node. A useful feature of the app is the ability to interact with the
visualisations. The tools on the right of the plot allow the user to examine
edges within a sequence/time slice, and query individual edges. The histograms
to the right provide summaries of edge spans over sequence and time.

@@

\clearpage \onecolumn \setcounter{figure}{0}  % Reset the figure counter
\renewcommand{\thefigure}{S\arabic{figure}}  % Change figure numbering to S1,
S2, etc. \section{Supplementary Material} \subsection{Simulation of truth
dataset (Fig. \ref{fig:Figure_2})} \label{subsec:sweep_simulation} Ancestral
histories of 300 samples were simulated with \texttt{SweepGenicSelection}
function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
in the recent past, a selective sweep was simulated with a beneficial allele
situated in the middle of a 5 Mb sequence. The frequency of the allele in the
population was set at 0.0001 at the beginning of the sweep. The allele fixed in
the population at a frequency of 0.9999. The strength of selection was set
using the selection coefficient, \textit{s = 0.25}. A time increment,
\textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
the ARG at a rate of 1e-6. For simulating history before occurrence of the
sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
used until coalescence was achieved.

>>>>

\subsection{User Interface} \label{subsec:User_Interface} \begin{figure}
\centering \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
\caption{Screenshot of the Edges tab in \texttt{tsbrowse}.
Visualisation of the edges in a simulated ARG with a strong selective
sweep in the middle of the genome (see Supplement for details).
Edges are shown in the main browser pane on the left, with additional
histograms summarising the edges on the right.
The effect of the sweep can be seen by the lack of edges crossing the 
centre of the simulated chromosome. Other tabs are accessed by the 
buttons on the top bar.
\label{fig:Figure_2}}
\end{figure}
% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
% visualisation pipelines even for large datasets. A key challenge with
% large-scale data visualisation is to maintain a responsive user interface.
% \texttt{Datashader} aggregates data and renders only the necessary data points
% at a given zoom level, thereby minimising the computational load on the
% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
% with individual data points, allowing users to gain deeper insights from the
% data. The \texttt{serve} step accepts an optional input file containing
% sequence annotations. This allows users to overlay contextual information about
% genes or other sequence features. 
The user interface is presented as a dashboard, with pages to
describe various aspects of the ARG. 
Figure~\ref{fig:Figure_2}
demonstrates the user interface using the Edges page as an example. 
The plot on the left is an overview of the 33,929 edges 
(parent-child relationships between nodes along the genome)
in a simulated ARG with a strong selective sweep (see
Sec.~\ref{subsec:sweep_simulation} for details).
Each edge is depicted as a horizontal line connecting the genomic coordinates of the
parent and child nodes on the x-axis, and the y-axis shows the time of 
either the parent or the child, as chosen by the user.
The user can interact with the main ``genome browser'' window
using a set of controls on the top-right corner (provided by Bokeh),
allowing the user to pan and zoom as required. The histograms to the
right then summarise the edges as depicted in the browser window.
A similar browser interface is provided for 
nodes
(e.g. Fig~\ref{fig:Supplementary_Figure_3}) 
and mutations 
(e.g. Fig~\ref{fig:Supplementary_Figure_4}).
\texttt{Tsbrowse} also provides an interactive table viewer with flexible
search and sorting utilities, which is a valuable debugging utility for 
developers.

@@

\clearpage \onecolumn \setcounter{figure}{0}  % Reset the figure counter
\renewcommand{\thefigure}{S\arabic{figure}}  
% Change figure numbering to S1, S2, etc.

\section{Supplementary Material} 
\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})} \label{subsec:sweep_simulation} 
Ancestral
histories of 300 samples were simulated with \texttt{SweepGenicSelection}
function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
in the recent past, a selective sweep was simulated with a beneficial allele
situated in the middle of a 5 Mb sequence. The frequency of the allele in the
population was set at 0.0001 at the beginning of the sweep. The allele fixed in
the population at a frequency of 0.9999. The strength of selection was set
using the selection coefficient, \textit{s = 0.25}. A time increment,
\textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
the ARG at a rate of 1e-6. For simulating history before occurrence of the
sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
used until coalescence was achieved.
